---
title: "589 project"
author: "all"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document: default
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

```{r}
library(sf)
library(spatstat.geom)
library(sp)
load("BC_Animal.Rda")
sapply(DATA, class)
str(DATA$Animal)

```

```{r}
plot(DATA$Window, main = "Spatial Boundaries of BC Parks")
suppressPackageStartupMessages(library(spatstat))
plot(DATA$Elevation,main="Elevation Map of BC Parks")
coords <- DATA$Animal[,c("X","Y")]
window_sf <- st_as_sf(DATA$Window)
window <- as.owin(window_sf)
animal_ppp <- ppp(x = coords$X,
                y = coords$Y,
                window = window,
                )
# Run plot() on the resulting ppp object.
plot(animal_ppp)
```

```{r}
fig <- persp(DATA$Elevation,
      theta = 10,
      phi = 30,
      expand = 20,
      border = NA,
      shade = 0.7,
      lphi = 10,
      r = 10,
      box = FALSE,
      apron = TRUE,
      colmap  =terrain.colors(100),
      visible = TRUE,
      main = ""
)
x <- DATA$Animal$X
y <- DATA$Animal$Y
# Overlay the park locations.
perspPoints(x = x, y = y,
            Z = DATA$Elevation, M = fig,
            pch = 16, col = "blue", cex = 0.6)
plot(cut(DATA$Elevation,
         5),col = terrain.colors(5),
     main = "Elevation classes")

points(DATA$Animal$X, DATA$Animal$Y, pch = 16, col = "blue", cex = 0.6)

b <- quantile(DATA$Elevation, probs = (0:5)/5, type = 2, na.rm = TRUE)
Zcut <- cut(DATA$Elevation, breaks = b)


V <- tess(image = Zcut)


xrange <- range(c(DATA$Animal$X, DATA$Elevation$xrange))
yrange <- range(c(DATA$Animal$Y, DATA$Elevation$yrange))
custom_win <- owin(xrange = xrange, yrange = yrange)

animal_ppp_full <- ppp(x = DATA$Animal$X,
                       y = DATA$Animal$Y,
                       window = custom_win)

counts <- quadratcount(animal_ppp_full, tess = V)

table_counts <- as.table(counts)
table_counts

```

```{r}

# Calculate the median elevation within British Columbia and at animal locations
median_bc <- median(DATA$Elevation$v, na.rm = TRUE)
elev_at_animals <- interp.im(DATA$Elevation, DATA$Animal$X, DATA$Animal$Y)
median_animals <- median(elev_at_animals, na.rm = TRUE)

cat("Median elevation in BC:", median_bc, "\n")
cat("Median elevation at animal locations:", median_animals, "\n")

# Generate KDE of elevation across BC
elev_all <- DATA$Elevation$v
elev_all <- elev_all[!is.na(elev_all)]

# Generate KDE of elevation at animal observation locations
elev_at_animals <- elev_at_animals[!is.na(elev_at_animals)]
kde_all <- density(elev_all)
kde_animals <- density(elev_at_animals)

# Plot both KDE distributions
plot(kde_all,
     main = "Elevation Distributions: BC vs. Animal Observations",
     xlab = "Elevation (meters)",
     ylab = "Density",
     col = "gray40", lwd = 2,
     xlim = range(c(kde_all$x, kde_animals$x)))
lines(kde_animals, col = "blue", lwd = 2)

# Add legend
legend("topright",
       legend = c("BC (All Pixels)", "Animal Observations"),
       col = c("gray40", "blue"),
       lwd = 2)

# Identify both the highest and lowest elevation animal observations
elev_at_animals_full <- interp.im(DATA$Elevation, DATA$Animal$X, DATA$Animal$Y)
max_index <- which.max(elev_at_animals_full)
min_index <- which.min(elev_at_animals_full)

highest_animal <- DATA$Animal[max_index, ]
lowest_animal  <- DATA$Animal[min_index, ]

cat("Highest elevation animal observation:\n")
print(highest_animal)

cat("Lowest elevation animal observation:\n")
print(lowest_animal)


```

```{r}
# Compute the spatial intensity (points per km²) for animal observations
library(spatstat.explore)
# For example, a 10x10 grid
nx <- 10
ny <- 10
# Count points in each quadrat
Q <- quadratcount(animal_ppp, nx=nx, ny=ny)
intensity(Q)
quadrat.test(Q)
pp.animal.km <- rescale(animal_ppp, s = 1000, unitname = "km")
area_km <- area.owin(Window(pp.animal.km))
n_animals <- npoints(animal_ppp)
intensity <- n_animals / area_km
intensity
plot(animal_ppp , pch = 16, cols = "black", cex = 0.6)
# Overlay the quadrat grid
plot(Q, add = TRUE, col = "red")
```

```{r}
# Plot quadrat-based intensity and overlay animal locations
plot(intensity(Q, image = TRUE))
plot(animal_ppp, pch = 16, cex = 0.3, col = "black", add = TRUE)

# Kernel density estimation using Diggle’s rule
kde <- density(animal_ppp, sigma = bw.ppl)

# Plot kernel density map with animal points
plot(kde, main = "Kernel Density of Animal Observation Intensity")
plot(animal_ppp, pch = 16, cex = 0.3, col = "black", add = TRUE)

```

```{r}
# Perform kernel intensity estimation with cross-validated bandwidth and identify hotspots for animal observations
bw_cv <- bw.ppl(animal_ppp)
kde_cv <- scanLRTS(animal_ppp, r = bw_cv)
plot(kde_cv, main = "Intensity Estimate (bw.ppl cross-validation)")
hotspot_mask <- eval.im(pchisq(kde_cv, df = 1, lower.tail = FALSE))
plot(Window(animal_ppp), add = TRUE, border = "darkgreen", lwd = 1)
plot(hotspot_mask)
plot(Window(animal_ppp), add = TRUE, border = "darkgreen", lwd = 1)

```

```{r}
# Estimate the relationship between animal intensity and elevation using rhohat
Zcut <- cut(DATA$Elevation, breaks = quantile(DATA$Elevation, probs = 0:4 / 4, type = 2))
animals <- unmark(animal_ppp)
quadratcount(animals, tess = tess(image = Zcut))
rho <- rhohat(animals, DATA$Elevation)
plot(rho)

```

```{r}
# Compute the inhomogeneous intensity surface for animal observations
lambda <- density(animal_ppp, sigma = bw.ppl, positive = TRUE)

# Generate Monte Carlo simulations for inhomogeneous K-function
en_K <- envelope(animal_ppp, Kinhom, simulate = expression(rpoispp(lambda)),
                 correction = "border", rank = 1, nsim = 19, fix.n = TRUE)

# Plot the K-function with simulation envelope
plot(en_K, main = "Ripley’s K-function with CSR Envelopes (Animal Observations)")


```

```{r}

```

## Forest

```{r}
# Visualize the Forest covariate image
plot(DATA$Forest, main = "Forest Cover (BC)", col = terrain.colors(100))

# Overlay animal observation locations
points(DATA$Animal$X, DATA$Animal$Y, pch = 16, col = "blue", cex = 0.5)


```

```{r}
# Estimate the relationship between animal intensity and forest cover
# Use rhohat to visualize conditional intensity of points vs. Forest
rho_forest <- rhohat(animal_ppp, covariate = DATA$Forest)

# Plot the result
plot(rho_forest, main = "Animal Intensity vs. Forest Cover",
     xlab = "Forest Cover", ylab = "Estimated Intensity")

```

**Animal intensity increases steadily with increasing forest cover.**

**The estimated intensity curve slopes upward, suggesting a positive correlation.**

**This indicates that animals are more likely to be observed in areas with higher forest density.**

```{r}
# Fit an inhomogeneous Poisson model using forest cover as a covariate
ppm_forest <- ppm(animal_ppp ~ forest, covariates = list(forest = DATA$Forest))

# Predict the fitted intensity surface based on the model
lambda_forest <- predict(ppm_forest)

# Generate simulation envelope for inhomogeneous K-function under fitted model
en_K_forest <- envelope(animal_ppp, Kinhom,
                        simulate = expression(rpoispp(lambda_forest)),
                        correction = "border",
                        rank = 1, nsim = 19, fix.n = TRUE)
plot(en_K_forest, main = "Inhomogeneous K-function (Fitted with Forest Cover)")


```

**After accounting for forest cover, there’s still extra clustering among animal observations, animals tend to be closer together than expected under a Poisson model fitted with forest cover.**

**Clustering happens at short to medium distances(0-200000), not over long distances**

```{r}
# Generate simulation envelope for homogeneous pair correlation (just for contrast)
pcf_hom_forest <- envelope(animal_ppp, pcf,
                            simulate = expression(rpoispp(lambda_forest)),
                            rank = 1, nsim = 19)
plot(pcf_hom_forest, main = "Pair Correlation Function (Homogeneous) - Forest Cover Model")

```

**The pair correlation function indicates that there’s residual small-scale clustering, suggesting additional spatial influences beyond forest cover.**

**The pair correlation function shows a spike at small distances that forest cover alone cannot explain.**

**This tells us:**

**Animals are clustering more closely together than expected, even within similar forest cover levels.**

**There's likely local-scale behavior or unobserved variables at play.**

**Possible causes:**

**Social behavior (e.g., herding, pack dynamics)**

**Shared resources (e.g., water, food patches)**

**Nesting/breeding areas**

**Missing covariates like elevation, proximity to water, or human presence**

```{r}
# Inhomogeneous pair correlation function under forest cover-fitted model
pcf_inhom_forest <- envelope(animal_ppp, pcfinhom,
                             simulate = expression(rpoispp(lambda_forest)),
                             nsim = 19, rank = 1)
plot(pcf_inhom_forest, main = "Inhomogeneous Pair Correlation (Fitted with Forest Cover)")
```

**The forest cover exhibits significant spatial clustering or aggregation at certain distances, as revealed by the inhomogeneous pair correlation analysis. The fitted functions help quantify the roles of different spatial processes in shaping the observed pattern.**

```{r}
# Extract all pixel values of Forest across BC
forest_all <- DATA$Forest$v
forest_all <- forest_all[!is.na(forest_all)]

# Extract Forest values at animal observation locations
forest_at_animals <- interp.im(DATA$Forest, DATA$Animal$X, DATA$Animal$Y)
forest_at_animals <- forest_at_animals[!is.na(forest_at_animals)]

# Compute KDEs
kde_forest_all <- density(forest_all)
kde_forest_animals <- density(forest_at_animals)

# Plot both KDEs
plot(kde_forest_all,
     main = "Forest Cover: BC vs. Animal Observations",
     xlab = "Forest Cover",
     ylab = "Density",
     col = "gray40", lwd = 2,
     xlim = range(c(kde_forest_all$x, kde_forest_animals$x)))
lines(kde_forest_animals, col = "blue", lwd = 2)

# Add legend
legend("topright",
       legend = c("BC (All Pixels)", "Animal Observations"),
       col = c("gray40", "blue"),
       lwd = 2)


```

**This is a nonparametric exploration of spatial covariate effects:**

**Animals are less likely to be observed in areas of:**

**Very low forest cover (urban, farmland, etc.)**

**Very high forest cover (possibly dense forest with poor visibility or accessibility)**

**Instead, observations are more common in moderate to moderately high forest cover, possibly representing edge habitats or accessible areas.**
