---
title: "589 project"
author: "all"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%OS')`"
output:
  html_document: default
---

```{r setup, include=FALSE}
# DO NOT ALTER CODE IN THIS CHUNK
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

```{r}
library(sf)
library(spatstat.geom)
library(sp)
load("BC_Animal.Rda")
sapply(DATA, class)
str(DATA$Animal)

```

```{r}
plot(DATA$Window, main = "Spatial Boundaries of BC")
suppressPackageStartupMessages(library(spatstat))
plot(DATA$Elevation,main="Elevation Map of BC Animals")
coords <- DATA$Animal[,c("X","Y")]
window_sf <- st_as_sf(DATA$Window)
window <- as.owin(window_sf)
animal_ppp <- ppp(x = coords$X,
                y = coords$Y,
                window = window,
                )
# Run plot() on the resulting ppp object.
plot(animal_ppp)
```

```{r}
fig <- persp(DATA$Elevation,
      theta = 10,
      phi = 30,
      expand = 20,
      border = NA,
      shade = 0.7,
      lphi = 10,
      r = 10,
      box = FALSE,
      apron = TRUE,
      colmap  =terrain.colors(100),
      visible = TRUE,
      main = ""
)
x <- DATA$Animal$X
y <- DATA$Animal$Y
# Overlay the park locations.
perspPoints(x = x, y = y,
            Z = DATA$Elevation, M = fig,
            pch = 16, col = "blue", cex = 0.6)
plot(cut(DATA$Elevation,
         5),col = terrain.colors(5),
     main = "Elevation classes")

points(DATA$Animal$X, DATA$Animal$Y, pch = 16, col = "blue", cex = 0.6)

b <- quantile(DATA$Elevation, probs = (0:5)/5, type = 2, na.rm = TRUE)
Zcut <- cut(DATA$Elevation, breaks = b)


V <- tess(image = Zcut)


xrange <- range(c(DATA$Animal$X, DATA$Elevation$xrange))
yrange <- range(c(DATA$Animal$Y, DATA$Elevation$yrange))
custom_win <- owin(xrange = xrange, yrange = yrange)

animal_ppp_full <- ppp(x = DATA$Animal$X,
                       y = DATA$Animal$Y,
                       window = custom_win)

counts <- quadratcount(animal_ppp_full, tess = V)

table_counts <- as.table(counts)
table_counts

```

```{r}

# Calculate the median elevation within British Columbia and at animal locations
median_bc <- median(DATA$Elevation$v, na.rm = TRUE)
elev_at_animals <- interp.im(DATA$Elevation, DATA$Animal$X, DATA$Animal$Y)
median_animals <- median(elev_at_animals, na.rm = TRUE)

cat("Median elevation in BC:", median_bc, "\n")
cat("Median elevation at animal locations:", median_animals, "\n")

# Generate KDE of elevation across BC
elev_all <- DATA$Elevation$v
elev_all <- elev_all[!is.na(elev_all)]

# Generate KDE of elevation at animal observation locations
elev_at_animals <- elev_at_animals[!is.na(elev_at_animals)]
kde_all <- density(elev_all)
kde_animals <- density(elev_at_animals)
y_max <- max(c(kde_all$y, kde_animals$y))
# Plot both KDE distributions
plot(kde_all,
     main = "Elevation Distributions: BC vs. Animal Observations",
     xlab = "Elevation (meters)",
     ylab = "Density",
     col = "gray40", lwd = 2,
     xlim = range(c(kde_all$x, kde_animals$x)),
     ylim = c(0, y_max))
lines(kde_animals, col = "blue", lwd = 2)

# Add legend
legend("topright",
       legend = c("BC (All Pixels)", "Animal Observations"),
       col = c("gray40", "blue"),
       lwd = 2)

# Identify both the highest and lowest elevation animal observations
elev_at_animals_full <- interp.im(DATA$Elevation, DATA$Animal$X, DATA$Animal$Y)
max_index <- which.max(elev_at_animals_full)
min_index <- which.min(elev_at_animals_full)

highest_animal <- DATA$Animal[max_index, ]
lowest_animal  <- DATA$Animal[min_index, ]

cat("Highest elevation animal observation:\n")
print(highest_animal)

cat("Lowest elevation animal observation:\n")
print(lowest_animal)


```

```{r}
# Compute the spatial intensity (points per km²) for animal observations
library(spatstat.explore)
# For example, a 10x10 grid
nx <- 10
ny <- 10
# Count points in each quadrat
Q <- quadratcount(animal_ppp, nx=nx, ny=ny)
intensity(Q)
quadrat.test(Q)
pp.animal.km <- rescale(animal_ppp, s = 1000, unitname = "km")
area_km <- area.owin(Window(pp.animal.km))
n_animals <- npoints(animal_ppp)
intensity <- n_animals / area_km
intensity
plot(animal_ppp , pch = 16, cols = "black", cex = 0.6)
# Overlay the quadrat grid
plot(Q, add = TRUE, col = "red")
```

```{r}
# Plot quadrat-based intensity and overlay animal locations
plot(intensity(Q, image = TRUE))
plot(animal_ppp, pch = 16, cex = 0.3, col = "black", add = TRUE)

# Kernel density estimation using Diggle’s rule
kde <- density(animal_ppp, sigma = bw.ppl)

# Plot kernel density map with animal points
plot(kde, main = "Kernel Density of Animal Observation Intensity")
plot(animal_ppp, pch = 16, cex = 0.3, col = "black", add = TRUE)

```

```{r}
# Perform kernel intensity estimation with cross-validated bandwidth and identify hotspots for animal observations
bw_cv <- bw.ppl(animal_ppp)
kde_cv <- scanLRTS(animal_ppp, r = bw_cv)
plot(kde_cv, main = "Intensity Estimate (bw.ppl cross-validation)")
hotspot_mask <- eval.im(pchisq(kde_cv, df = 1, lower.tail = FALSE))
plot(Window(animal_ppp), add = TRUE, border = "darkgreen", lwd = 1)
plot(hotspot_mask)
plot(Window(animal_ppp), add = TRUE, border = "darkgreen", lwd = 1)

```

```{r}
# Estimate the relationship between animal intensity and elevation using rhohat
Zcut <- cut(DATA$Elevation, breaks = quantile(DATA$Elevation, probs = 0:6 / 6, type = 2))
animals <- unmark(animal_ppp)
quadratcount(animals, tess = tess(image = Zcut))
rho <- rhohat(animals, DATA$Elevation)
plot(rho,xlim=(c(0,3000)))

```

```{r}
# Compute the inhomogeneous intensity surface for animal observations
lambda <- density(animal_ppp, sigma = bw.ppl, positive = TRUE)

# Generate Monte Carlo simulations for inhomogeneous K-function
en_K <- envelope(animal_ppp, Kinhom, simulate = expression(rpoispp(lambda)),
                 correction = "border", rank = 1, nsim = 19, fix.n = TRUE)

# Plot the K-function with simulation envelope
plot(en_K, main = "Ripley’s K-function with CSR Envelopes (Animal Observations)")


```

The observed inhomogeneous K-function closely follows the expected K-function under complete spatial randomness (CSR), indicating no significant spatial clustering or inhibition beyond what is explained by the fitted intensity model.

```{r}
# Compute intensity surface for animal observations
lambda <- density(animal_ppp, sigma = bw.ppl(animal_ppp), positive = TRUE)

# Generate simulation envelope for homogeneous pair correlation function
pcf_hom <- envelope(animal_ppp, pcf, simulate = expression(rpoispp(lambda)),
                    rank = 1, nsim = 19)

# Plot the homogeneous pair correlation function with envelope
plot(pcf_hom, main = "Pair Correlation Function (Homogeneous) - Animal Observations")
pcf_inhom <- envelope(animal_ppp, pcfinhom, nsim = 19,
                      simulate = expression(rpoispp(lambda)), rank = 1)

# Plot the inhomogeneous pair correlation function with simulation envelope
plot(pcf_inhom, main = "Inhomogeneous Pair Correlation Function (Animal Observations)")
```

observations show **significant small‑scale clustering**—many individuals occur within short distances of each other—while exhibiting no detectable interaction at larger scales.

```{r}
animal <- unmark(animal_ppp)

elev_rho <- rhohat(animal, DATA$Elevation)
# Plot with proper axis limits (no negative elevation)
plot(elev_rho,main = "Estimated Intensity vs Elevation", 
     xlab = "Elevation (m)",xlim=(c(0,3000)))

# Estimate intensity (rho) as a function of forest cover
forest_rho <- rhohat(animal, DATA$Forest)

# Plot the intensity as a function of forest cover
plot(forest_rho, xlab = "Forest Cover (%)",
     ylab = "Estimated Intensity")
```

```{r}
hfi_rho <- rhohat(animal, DATA$HFI)
plot(hfi_rho)
dis_rho <- rhohat(animal, DATA$Dist_Water)
plot(dis_rho)
```

```{r}
elev <- DATA$Elevation
cover <- DATA$Forest
hfi <- DATA$HFI
dis <- DATA$Dist_Water
cor.im(elev,cover,hfi,dis,use ="pairwise.complete.obs")
```

```{r}
animal_unmarked <- unmark(animal_ppp)
model <- ppm(animal_unmarked  ~ elev + I(elev^2) + cover + I(cover^2)+hfi+I(hfi^2)+dis+I(dis^2),covariates = list(elev = elev, cover = cover,hfi=hfi,dis=dis))
model
cat(AIC(model))
plot(model,se=FALSE,superimpose=FALSE)
plot(animal,pch=16,cex=0.6,color="white",add=TRUE)

model1_1 <- ppm(animal_unmarked  ~ hfi+I(hfi^2),covariates = DATA)
cat(AIC(model1_1))
model2_1 <- ppm(animal_unmarked  ~hfi+I(hfi^2)+ elev + I(elev^2) , covariates = DATA)
cat(AIC(model2_1))
model3_1 <- ppm(animal_unmarked  ~hfi+I(hfi^2)+ elev + I(elev^2) +cover + I(cover^2),covariates = DATA)
cat(AIC(model3_1))

```

```{r}
quadrat.test(model)

```

```{r}
null_model <- ppm(animal_unmarked ~ 1)
null_model
AIC(model):AIC(null_model)
AIC(null_model)-AIC(model)
```

```{r}
log_intensity <- log(predict(model, type = "trend", n = 500))  
# Plot the log-intensity surface
plot(log_intensity, main = "Log Fitted Intensity of Parks")
plot(animal, pch = 16, cex = 0.6, col = "white", add = TRUE)
# Median forest cover and elevation
E_cover <- median(cover[], na.rm = TRUE)
E_elev <- median(elev[], na.rm = TRUE)
E_hfi <- median(hfi[], na.rm = TRUE)
E_dis <- median(dis[], na.rm = TRUE)
elev_effect <- effectfun(model, "elev", 
                         cover = E_cover, hfi = E_hfi, dis = E_dis, 
                         se.fit = FALSE)

cover_effect <- effectfun(model, "cover", 
                          elev = E_elev, hfi = E_hfi, dis = E_dis, 
                          se.fit = FALSE)

hfi_effect <- effectfun(model, "hfi", 
                        elev = E_elev, cover = E_cover, dis = E_dis, 
                        se.fit = FALSE)

dis_effect <- effectfun(model, "dis", 
                        elev = E_elev, cover = E_cover, hfi = E_hfi, 
                        se.fit = FALSE)

# 📉 Elevation effect
plot(elev_effect, legend = FALSE,
     main = paste("Effect of Elevation\n(at forest cover =", round(E_cover, 1), "%)"),
     xlab = "Elevation (m)",
     ylab = "log(lambda)",
     xlim = c(-130, 3000))

# 🌲 Forest cover effect
plot(cover_effect, legend = FALSE,
     main = paste("Effect of Forest Cover\n(at elevation =", round(E_elev, 1), "m)"),
     xlab = "Forest Cover (%)",
     ylab = "log(lambda)")

# 👣 HFI effect
plot(hfi_effect, legend = FALSE,
     main = paste("Effect of Human Footprint Index\n(at HFI =", round(E_hfi, 1), ")"),
     xlab = "Human Footprint Index",
     ylab = "log(lambda)")

# 📏 Distance effect
plot(dis_effect, legend = FALSE,
     main = paste("Effect of Distance to Road\n(at Distance =", round(E_dis, 1), ")"),
     xlab = "Distance to Road (km)",
     ylab = "log(lambda)")

```

```{r, cache=TRUE, warning=FALSE}
# Visualize the Dist_Water covariate image
plot(DATA$Dist_Water, main = "Distance to Water (BC)", col = terrain.colors(100))

# Overlay animal observation locations
points(DATA$Animal$X, DATA$Animal$Y, pch = 16, col = "blue", cex = 0.5)

# Estimate the relationship between animal intensity and distance to water
# Use rhohat to visualize conditional intensity of points vs. Dist_Water
rho_dist <- rhohat(animal_ppp, covariate = DATA$Dist_Water)

# Plot the result
plot(rho_dist, main = "Animal Intensity vs. Distance to Water",
     xlab = "Distance to Water", ylab = "Estimated Intensity")

# Fit an inhomogeneous Poisson model using distance to water as a covariate
ppm_dist <- ppm(animal_ppp ~ dist, covariates = list(dist = DATA$Dist_Water))

# Predict the fitted intensity surface based on the model
lambda_dist <- predict(ppm_dist)

plot(log(lambda_dist),
     main = "Log-Intensity of Animal Occurrences (Distance to Water)")

# Overlay the observed animal locations
plot(animal_ppp, pch = 16, cex = 0.3, col = "black", add = TRUE)
# Generate simulation envelope for inhomogeneous K-function under fitted model
en_K_dist <- envelope(animal_ppp, Kinhom,
                      simulate = expression(rpoispp(lambda_dist)),
                      correction = "border",
                      rank = 1, nsim = 19, fix.n = TRUE)

# Plot the inhomogeneous K-function with envelope
plot(en_K_dist, main = "Inhomogeneous K-function (Fitted with Dist_Water)")


```

## Dist_Water plot with overlay animal points

Most animals are observed **closer to water sources**.

**Central & Southern BC have particularly high densities of points**:

This may reflect ecological preference or simply **sampling bias** (e.g., more data collected in accessible regions).

The plot **suggests a negative correlation** between **distance to water** and **animal presence**.

## Animal intensity vs dist

**Squirrel observations are most intense within \~0–2,000 meters** of water sources.

There are **local peaks** around 5,000 m and 10,000 m, suggesting: Either suitable squirrel habitats also exist at moderate distances from water, or the landscape (e.g., elevation, forest patches) in those areas still provides viable conditions. After \~13,000 m, the estimated intensity drops sharply to near zero, indicating that this species is rarely found far from water.

## Log intensity plot

This squirrel species is more likely to occur near water.\
Observed points cluster well with the yellow/orange regions

## **inhomogeneous K-function** with simulation envelopes

Kinhom(r) estimates the expected number of other points within distance r of a typical point, after adjusting for Dist_Water

Observed line (black) slightly exceeds the simulation envelope at small to medium r:

-   This suggests residual clustering of squirrel observations not fully explained by Dist_Water alone.

-   It means: even after accounting for water proximity, squirrels tend to occur closer together than expected under a Poisson model driven only by Dist_Water.

At large r (beyond \~200,000 m):

-   The observed curve converges with the simulated envelope.

-   This means the model performs reasonably well at capturing large-scale spatial structure.

-   This result could also be ignored because rarely do we see a location that is over 200,000 meters away from water while having squirrels living there.

There may be additional spatial drivers influencing the point pattern.

```{r, warning=FALSE}
# Generate simulation envelope for homogeneous pair correlation (just for contrast)
pcf_hom_dist <- envelope(animal_ppp, pcf,
                         simulate = expression(rpoispp(lambda_dist)),
                         rank = 1, nsim = 19)

plot(pcf_hom_dist, main = "Pair Correlation Function (Homogeneous) - Dist_Water Model")

# Inhomogeneous pair correlation function under dist_water-fitted model
pcf_inhom_dist <- envelope(
  animal_ppp,
  fun = pcfinhom,
  lambda = lambda_dist, 
  simulate = expression(rpoispp(lambda_dist)),
  nsim = 19,
  rank = 1
)

plot(pcf_inhom_dist, main = "Inhomogeneous Pair Correlation (Fitted with Dist_Water)")

```

## **homogeneous PCF** plot

**Strong clustering at small distances**:

-   The observed PCF rises sharply to values **above 30** at short distances.

-   This indicates **very strong local clustering**, far exceeding what the model based on `Dist_Water` explains.

Suggests that assuming homogeneity (i.e., ignoring spatial covariates like `Dist_Water`) is not realistic for this data.

## **inhomogeneous PCF** plot

Sharp spike at small distances (r \< \~20,000 m), Beyond \~50,000 m, the observed curve aligns with the envelope, at broader scales, the Dist_Water covariate adequately captures spatial variation in intensity.

**`Dist_Water` alone cannot explain fine-scale clustering** of the squirrels.

```{r}
# Extract all pixel values of Dist_Water across BC
dist_all <- DATA$Dist_Water$v
dist_all <- dist_all[!is.na(dist_all)]

# Extract Dist_Water values at animal observation locations
dist_at_animals <- interp.im(DATA$Dist_Water, DATA$Animal$X, DATA$Animal$Y)
dist_at_animals <- dist_at_animals[!is.na(dist_at_animals)]

# Compute KDEs
kde_dist_all <- density(dist_all)
kde_dist_animals <- density(dist_at_animals)

# Plot both KDEs
plot(kde_dist_all,
     main = "Distance to Water: BC vs. Animal Observations",
     xlab = "Distance to Water",
     ylab = "Density",
     col = "gray40", lwd = 2,
     xlim = range(c(kde_dist_all$x, kde_dist_animals$x)))
lines(kde_dist_animals, col = "blue", lwd = 2)


# Add legend
legend("topright",
       legend = c("BC (All Pixels)", "Animal Observations"),
       col = c("gray40", "blue"),
       lwd = 2)

```

## KDE plot

The squirrels appears to **prefer habitats within a few kilometers of water**, far more than would occur by chance.

This supports our earlier finding from `rhohat()`, the inhomogeneous K-function, and our intensity map.
